---
title: "dmfind tutorial"
author: "Ettore Mosca"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network diffusion-based analysis of omics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
devtools::load_all(".")
```

##Introduction
The R package dmfind implements the algorithms described in @Bersanelli2016. In this tutorial we will use igraph package for preparing the input data and plotting the results.

##Installation
The package can be installed from within R:
```{r, eval=FALSE}
library(devtools)
install_github("emosca-cnr/dmfind002", build_vignettes = TRUE)
library(dmfind)
```

##Input data
The package requires an adjacency matrix and one or more input vectors. Let's create a simulated input dataset, including: a normalized adjacency matrix (A), a two-columns matrix (D) with two vectors representing descriptive statistics of two sample groups, and a binary vector where positive elements indicates the significant elements of an inferential statistics (u):

```{r, eval=TRUE, message=FALSE}
library(igraph)

edge_list_ig <- list(c(2, 3, 4, 12), c(1,3,4,5, 13), c(1, 2, 3), c(1,2,5), c(2,4, 7, 9), 7,
                              c(5,6,8), c(7,10), c(5,10), c(8,9,11), 10, 1, 2)
edge_list <- do.call(rbind, lapply(1:length(edge_list_ig),
                                   function(i) cbind(i, edge_list_ig[[i]])))
G <- graph_from_adj_list(edge_list_ig, 'all')
N <- length(V(G))
V(G)$name <- 1:N
A <- dmfind::normalize_adj_mat(as.matrix(as_adjacency_matrix(G)))
D <- matrix(abs(c(rnorm(N, 0, 0.2), rnorm(4, 2), rnorm(N-4, 0, 0.2))), ncol = 2,
            dimnames = list(1:N, 1:2))
temp_tgt <- D[1:4, 2]
temp_src_idx <- order(-D[, 2])[1:4]
temp_src <- D[temp_src_idx, 2]
D[1:4, 2] <- temp_src
D[temp_src_idx, 2] <- temp_tgt
u <- matrix(c(rep(1, 4), rep(0, N-4)), ncol=1, dimnames = list(1:N))
print(D)
print(u)
```
Note that elements ${1, 2, 3, 4}$ are enriched in information in `D[, 2]` in comparison to `D[, 1]`, and are positive in $u$. At this point we can calculate the network smoothing index $(S)$ of each sample group and its variation $\Delta S$:
```{r, eval=TRUE, message=FALSE, eval=T}
dS_Ds <- dmfind::calc_dS(D, A, classes = c(1, 2), eps = 4)
print(dS_Ds)
plot(dS_Ds$dX0, dS_Ds$dS, xlab='dX0', ylab='dS', pch='')
text(dS_Ds$dX0, dS_Ds$dS, rownames(dS_Ds), cex=0.7)
```

In case of the vector $u$ we calculate the permutation-adjusted network smoothing index $(Sp)$:

```{r, eval=TRUE, message=FALSE}
Sp_u <- dmfind::calc_Sp(u, A, k=9, eps = 4)
print(Sp_u)
```

Network resampling allows to assess the list ranked by $\Delta S$ (or $Sp$) for the presence of significantly connected components:
```{r, eval=TRUE, message=FALSE}
res <- dmfind::NR(edge_list, sort(array(dS_Ds$dS, dimnames = list(rownames(dS_Ds))), decreasing = TRUE), k=99)
print(res$NR_summary)
plot(log10(res$NR_summary$p), xlab='rank', type='l', ylab="log10(p)", ylim=c(-2,0))
```

Let's plot the network underlying the region with the highest $\Delta S$:
```{r, eval=TRUE, message=FALSE}
V(G)$color <- heat.colors(16)[
  dmfind::round(
    linear_map(
      res$NR_summary$ranking_score[match(V(G)$name, names(res$NR_summary$ranking_score))], 1, 16))]
par(mar=c(0, 0, 0, 0))
plot.igraph(G, vertex.size=10, vertex.label.cex=1)
```

##Reference

---
references:
- id: Bersanelli2016
  title: Network diffusion-based analysis of high-throughput data for the detection of differentially enriched modules
  author:
  - family: Bersanelli
    given: Matteo
  - family: Mosca
    given: Ettore
  - family: Remondini
    given: Daniel
  - family: Castellani
    given: Gastone
  - family: Milanesi
    given: Luciano
  container-title: Scientific Reports
  volume: 6
  DOI: 10.1038/srep34841
  page: 34841
  type: article-journal
  issued:
    year: 2016
---
